Leadership across cultures and contexts: an exploratory analysis

### 2020-04-30 leadership_across_cultures_contexts.Rmd
Many edits to intro, methods, results.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning=FALSE, message=F)
library(pander)
library(patchwork)
library(gt)
loadd(all_study_vars)
loadd(df_qual)
loadd(df_func)
loadd(plot_func_qual)
loadd(qualities_support_txt)
loadd(qualities_support_cult)
loadd(functions_support_txt)
loadd(functions_support_cult)
loadd(highstatus_plot)
loadd(plot_costs_benefits)
loadd(plot_pubdate)
loadd(m_lpca_qual)
loadd(m_lpca_func)
source("R/cheapobjects.R")
```

@@

Research on leadership within anthropology extends to the birth of the discipline. Early on, anthropologists (and some sociologists) realized that leadership and followership were critical to understanding human psychology, social organization, and culture [e.g., @mumford_origins_1906; @firth1927economic; @morgan_ancient_1877; @myres_influence_1917]. Anthropologists documented some sort of leadership among every ethnographically observed society [@brown_human_1991; @lewis_leaders_1974], and across many social contexts, such as families and kin groups [@dussart2000politics], ritual [@singh_cultural_2017], work groups [@macfarlan_competitive_2012], conflicts between groups [@glowacki_evolutionary_2017], and in nonindustrial political groups [@cohen_comparative_1967]. For review, see @garfield_review_2019.

Despite decades of research and theorizing about leadership in numerous social science disciplines, to our knowledge there is no comprehensive or systematic investigation of leader functions and qualities, and the costs and benefits of leadership and followership, across a representative sample of nonindustrial societies. Moreover, outside of anthropology, most research and theorizing on leadership is based on data from Western or Westernized, postindustrial societies [@von_rueden_leadership_2015].

@@

There is now well established consensus within the behavioral sciences on the importance of data from the full range of human cultural diversity. Systemic biases in much of social science are well known [@henrich_weirdest_2010, *The weirdest people in the world?*], and might be due, in part, to the lack of data from populations other than American undergaduates. @henrich_weirdest_2010 in their initial critique of the over-realiance on WEIRD samples emphasized a holistic approach to cultural diversity within and across "telescoping" levels of human societies.

@@

Anthropology, on the other hand, has been criticized for exoticizing and essentializing the "Other" [@said1979; @abu2008]. @abu2008 [p. 470], for instance, notes that

@@

There is as much diversity within non-Western cultures as there is within Western ones, and there is also substantial overlap of the behaviors of people everywhere [@brown_human_1991].

Extending the study of leadership to include evolutionary theory and data from diverse, small-scale and traditional societies (i.e., non-WEIRD) can facilitate more generalizable theories and offer applications to shape organizational structure across populations (i.e., WEIRD settings) in ways which are more consistent and less mismatched with human evolutionary history and evolved leader-follower psychologies [@von_rueden_leadership_2015; @garfield_review_2019; @van_vugt_selected_2010]. But such approaches are also at risk of failing to incorporate the message of the larger WEIRD people problem: diversity of cultural, social, and ecological contexts can be strong divers of human behavior and group dynamics.

Leaders substantially shape group structures and outcomes across cultures. The structure of groups and their goals, in turn, as well as their subsistence strategies, can favor particular leader qualities.

as well as the costs and benefits to individuals involved. Cultural norms, capacities of competing leaders, inter-group dynamics, and follower preferences can also impact leader agendas. The interactions between leader strategies, pressures of followership, and group dynamics are likely to strongly contribute to organizational variation across human populations including social and economic trajectories [@wiessner_risk_1982] and cultural and biological evolutionary processes [@henrich_big_2015].

<!-- NOT SURE WHERE THIS SENTANCE SHOULD GO OR COULD BE CUT -->
Within every population there are particular contexts in which one or a few individuals maintain a disproportionate level of influence in the group, function to facilitate group goals, and provides services to the group.

@@

Evolutionary theoretical models and empirical studies often seek to integrate findings within a broad framework of human cultural diversity. The universality of human leadership systems and the incredible diversity among the qualities of leaders across cultures and contexts and the functions leaders serve is difficult to capture. Scholars are then left to haphazardly self-select ethnographic cases or rely on the self-selected reviews of others for generalizations [e.g., @garfield_review_2019; @von_rueden_leadership_2015]. Here, we seek to provide the first empirical and near comprehensive view of the qualities and functions of leaders across human societies from a representative cross-cultural sample. We also provide the first comparative view describing the costs and benefits of leadership for both leaders and followers in an evolutionary biological framework. We also aim to explore how the qualities and functions of leaders vary across social contexts and levels of cultural complexity

In sum, we seek to empirically and systematically accomplish the following four goals: 1) identify the most cross-culturally frequent *qualities of leaders* and the *functions they serve*. 2) Identify *features* of the qualities and functions of leaders (e.g., do particular variables tended to covary or cluster together?) in reference to theoretical models or novel interpretations. Such features might lend support to existing theories; for instance they might correspond to qualities associated with "prestige" or "dominance" [@henrich_evolution_2001] or functions associated with conflict or cooperation [@gavrilets_collective_2015; @glowacki_evolutionary_2017; @hooper_theory_2010]. 3) Identify the most cross-culturally frequent *costs and benefits* of leadership for both leaders and followers. Such exploratory results stand to facilitate more fine-grained approaches within leadership studies and greater theoretical and empirical work beyond community leadership, while also revealing the importance of a broad, cross-cultural perspective in understanding the psychology and cultural patterning of perhaps the most foundational component of human social behavior and organization.

@@

We estimate the cross-cultural frequency in the functions and qualities of leaders and the costs and benefits of leadership for both leaders and followers using an ethnographic database previously built and published (doi.org/10.5281/zenodo.2541999), from which @garfield_evolutionary_2019 "tested" three theoretical models against the ethnographic record. The electronic Human Relations Area Files (eHRAF) is a digital collection of ethnographies, subject-coded at the paragraph level following the Outline of Cultural Materials (OCM) coding scheme. The eHRAF is a subset of the larger HRAF and is continually growing in scope.

Whereas @garfield_evolutionary_2019, documented supporting evidence for measures of leadership in reference to influential theoretical models, our goals here are to let the ethnographic record "speak for itself" in understanding variation in the dimensions of leadership we are interested in, i.e., our methods and analyses are exploratory and not strictly theoretically driven. We draw on the same database of `r nrow(all_data)` ethnographic paragraphs (termed text records) extracted from the 60-culture Probability Sample Files (PSF) of the eHRAF database [@naroll_proposed_1967]. The PSF aims to mitigate Galton's problem by randomly sampling of one culture with high quality ethnographic coverage from each of 60 culture regions. It comprises over `r trunc(sum(leader_cult['ehraf_pages'])/1000)`K pages of digitized text.

@@

The operationalized qualities and functions (See Table \@ref(tab:qualfuncvars)) were developed from the ethnographic text records in the database. First ZG read a subset of text records and developed initial variable lists and operationalizations of the qualities and functions in those text records. Then, KS coded all `r nrow(all_data)` text records (in a random order) for the operationalized qualities, functions, cost, and benefits, while also creating new variables and operationalizations for additional qualities and functions encountered. After coding half of the random sample KS and ZG revised the operational definitions to produce the final list of variables in Table \@ref(tab:qualfuncvars). ZG then followed the same process of coding the entire sample of `r nrow(all_data)` text records for the `r length(c(quality_vars, function_vars, leader_cost_vars, leader_benefit_vars, follower_benefit_vars, follower_cost_vars))` operationalized variables, as well as a measure of group structure coded as: *residential subgroup*, which includes informal groups of co-residents, age-based groups, performance groups; *kin group*, which includes lineages, phraties, and other kin-based organizations; *economic group*, which includes subsistence groups, market groups, and other groups with primarily economic goals; *military group*, which includes all groups related to inter-group conflict; *religious group*, which includes groups collected for spiritual or supernatural purpose; *political group (community)*, which includes political groups at the level of the community, i.e., the political leader can potentially directly interact with most community members such as most village headmen, or *political group (supracommunity)*, which includes political groups that encompass multiple residential communities, such as complex chiefdoms, regional political leaders, and kings or state-level leaders. ZG and KS then discussed every instance of disagreement between their respective coding to produce a consensus matrix, which is used in the analyses here and added to the published data repository (a data frame named *leader_text2*).

<!-- DISCUSS IRR. -->

```{r qualfuncvars, echo=FALSE, message=F, warning=FALSE}
library(readr)
qual_func_var_tab<-read_csv('func_qual_definitions.csv', na = 'NA')
library(knitr)
kable(qual_func_var_tab, caption = "Operational definitions for coded leader qualities and functions.")
```

Coding operationalizations of the costs and benefits of leadership were developed *a priori* based on our reading of the evolutionary theoretical literature from anthropology and biology. For both followers and leaders, we code if there is evidence of costs paid or benefits received in the following categories: social status, reduced or increased risk of harm or conflict, social services, territory, mating, fitness, food resources, material resources, and other miscellaneous costs or benefits not captured by these categories. See Table \@ref(tab:costbenefitvars). There are multiple instances of conceptual overlap between many of our researcher-coded variables, in particular between various qualities and related costs or benefits. The operationalization of costs and benefits generally capture a wider range of evidence than related qualities and focus more on evidence of transaction rather than description. We return to this point in the limitations section.

```{r costbenefitvars, echo=FALSE, message=F, warning=FALSE}
library(readr)
cost_benefit_vars_tab<-read_csv('costs_benefits.csv', na = 'NA')
library(knitr)
kable(cost_benefit_vars_tab, caption = "Operational definitions for coded benefits and costs for leaders and followers.")
```

We illustrate our coding with an example from Andamanese hunter-gatherers [@az02-001, p. 45]:

@@

This text record was coded has having evidence for the functions *Resolve conflict*, *Organize cooperation*, *Movement/migration*, *Provide subsistence*, and *Misc. social functions*. For qualities, there was evidence for *High status*. Although authority is mentioned, we did not code evidence for this variable because the chief "possesses no power to punish or enforce obedience to his wishes," contrary to our operationalization. For additional details on ethnographic sampling and creation of the *leadershipdata* package see @garfield_evolutionary_2019.

@@

We analyzed data from our researcher-coded operationalized variables at the text-record level and at the culture-level. Our first goal was to assess the proportion of text records and cultures that provided evidence for each of our coded variables to discover the qualities, functions, benefits and costs of leadership that are most and least frequently noted in the ethnographic record. We computed confidence intervals for the proportion of text records supporting each variable using generalized linear mixed effects models and random intercepts based on document authors nested within cultures (which partially accounts for potential variation due to differences in ethnographers such as gender, nationality, or theoretical framework). We computed confidence intervals for the proportion of cultures with evidence for each variable using a cluster bootstrap.

@@

Our third goal was to determine which leader and follower benefits and costs were most commonly mentioned in the ethnographic record, computing 95% CIs as above and to estimate which group structure types were associated with the most overall benefits and costs for leaders.

@@

The geographic distribution of the cultures in our sample is displayed in Figure \@ref(fig:geographic), with cultures labeled by subsistence strategy and represented by a symbol proportional to the number of documents that provided ethnographic texts for that culture. Our search yielded `r nrow(all_data)` text records from `r length(unique(leadershipdata::documents$d_ID))` unique documents describing `r length(unique(all_data$d_culture))` cultures. The number of text records per culture ranged from `r min(leader_cult$number_leader_records, na.rm=T)` to `r max(leader_cult$number_leader_records, na.rm=T)`, with a median of `r median(leader_cult$number_leader_records, na.rm=T)`, a mean of `r signif(mean(leader_cult$number_leader_records, na.rm=T), 3)`, and a standard deviation of `r signif(sd(leader_cult$number_leader_records, na.rm=T), 3)`. The text records were generally short, ranging from `r textstats$min` to `r textstats$max` words, with a median word count of `r textstats$median`, a mean of `r textstats$mean`, and standard deviation of `r textstats$sd`. Of the `r nrow(all_data)` text records, `r final_record_count` had evidence for at least one variable in the study.

@@

There was a weak positive relationship between the number of leader text records we retrieved from each culture and the total number of pages of ethnography for each culture in eHRAF (r = `r signif(cor_pages_tr$estimate, 2)`, p = `r signif(cor_pages_tr$p.value, 2)`). A loess regression suggested that the number of records initially increased with the total number of pages of ethnography, plateauing at about 35 records around 4500 pages (see Figure \@ref(fig:eHRAFpages)).

@@

The type of group to which a leader text referred was coded into seven researcher-coded categories: residential groups, kin groups, economic groups, political groups at the community level, political group beyond the community level, military groups, and religious groups (see Methods section). Figure \@ref(fig:grouptype) indicates the proportion of text records that were categorized into each of these groups, broken out by subsistence strategy. We rely on a modified classification of subsistence from the HRAF, collapsing “hunter-gatherers” and “primarily hunter-gatherers” into a single category, “hunter-gatherers,” and by collapsing “agro-pastoralists” and “pastoralists” into a single “pastoralists” category. We renamed the HRAF’s “Other” subsistence category, which is applied to societies with various combinations of hunting, fishing, gathering, pastoralism, and horticulture or intensive agriculture, as “mixed.” This produced a subsistence classification with five groups: Hunter-gatherers, Pastoralists, Horticulturalists, Agriculturalists, and Mixed.

@@

@@

As Figure \@ref(fig:grouptype) reveals, the distribution of text records by group type varies across subsistence type. Across all subsistence types, the most frequent group type is political groups at the community level, with the exception of agriculturalists in which supracommunity political groups are the most frequent group type. Leadership beyond the community level is relatively rare outside of agricultural populations and is especially rare among hunter-gatherers. Leadership within residential groups is most frequently associated with hunter-gatherers (`r hg_residential_pct`% of the total sample of hunter-gatherer text records) and leadership within kin groups is most frequently associated with hunter-gatherers (`r hg_kin_pct`% of the total hunter-gatherers sample) and horticulturalists (`r hort_kin_pct`% of the total horticulturalists sample).

@@

@@

For each of the `r length(c(quality_vars, function_vars, leader_cost_vars, leader_benefit_vars, follower_benefit_vars, follower_cost_vars))` coded variables (see Tables \@ref(tab:qualfuncvars) and \@ref(tab:costbenefitvars)), we computed the percentage of text records that provided support, and we computed the percentage of cultures with at least one text record providing support for that variable. See Figure \@ref(fig:varsupport).

@@

We then investigated if the proportion of text records with evidence for high status differed among the different levels of social and political organization. There were modest differences. Somewhat counter-intuitively, the proportion of text records indicating high status was lowest at the highest level of political organization, perhaps because it would have been superfluous to note that kings and chiefs were high status. See Figure \@ref(fig:groupstatus).

```{r groupstatus, fig.width=10, fig.cap="Estimated marginal means of a mixed effects logistic regression model of evidence for high status (0/1) as a function of group type, with random effects for ethnographer nested within culture."}
((plot_emm_status_group2 / plot_emm_status_sex2) | (plot_emm_status_subsistence2 / plot_emm_status_region2)) + plot_annotation(title = "Predictors of high status", tag_levels = 'A')
```

Finally, we examined the relationship between evidence for high status and subsistence strategy and continental region. See Figure \@ref(fig:statusregion).

```{r statusregion, fig.cap="Estimated marginal means of a mixed effects logistic regression model of evidence for high status (0/1) as a function of subsistence strategy and region as covariates, with random effects for ethnographer nested within culture."}
```

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning=FALSE, message=F)
library(readr)
library(knitr)
library(pander)
library(patchwork)
library(gt)
loadd(all_study_vars)
loadd(df_qual)
loadd(df_func)
loadd(plot_func_qual)
loadd(qualities_support_txt)
loadd(qualities_support_cult)
loadd(functions_support_txt)
loadd(functions_support_cult)
loadd(highstatus_plot)
loadd(plot_costs_benefits)
loadd(plot_pubdate)
loadd(m_lpca_qual)
loadd(m_lpca_func)
source("R/cheapobjects.R")
```

@@

Research on leadership within anthropology extends to the birth of the discipline. Early on, anthropologists (and some sociologists) realized that leadership and followership were critical to understanding human psychology, social organization, and culture [e.g., @mumford_origins_1906; @firth1927economic; @morgan_ancient_1877; @myres_influence_1917]. Anthropologists have documented some sort of leadership among every ethnographically observed society [@brown_human_1991; @lewis_leaders_1974], and across many social contexts, such as families and kin groups [@dussart2000politics], ritual [@singh_cultural_2017], work groups [@macfarlan_competitive_2012], conflicts between groups [@glowacki_evolutionary_2017], and in nonindustrial political groups [@cohen_comparative_1967]. For review, see @garfield_review_2019.

Leaders substantially shape group structures and outcomes across cultures. The structure of groups and their goals, in turn, as well as their subsistence strategies, can favor particular leader qualities. Despite decades of research and theorizing about leadership in numerous social science disciplines, however, to our knowledge there is no comprehensive or systematic investigation of leader functions and qualities, and the costs and benefits of leadership and followership, across a representative sample of nonindustrial societies. Moreover, outside of anthropology, most research and theorizing on leadership is based on data from Western or Westernized, postindustrial societies [@von_rueden_leadership_2015].

@@

There is now well established consensus within the behavioral sciences on the importance of data from the full range of human cultural diversity. Systemic biases in much of social science are well known, and might be due, in part, to the lack of data from populations other than American undergaduates [@henrich_weirdest_2010, *The weirdest people in the world?*].

@@

Anthropology, on the other hand, has been criticized for exoticizing and essentializing the "Other" as part of its colonial history [@said1979; @abu2008]. @abu2008 [p. 470], for instance, notes that

@@

There is as much diversity within non-Western cultures as there is within Western ones, and there is also substantial overlap of the behaviors of people everywhere [@brown_human_1991]. @henrich_weirdest_2010 in their critique of the over-realiance on WEIRD samples emphasized a holistic approach to cultural diversity within and across "telescoping" levels of human societies. It is therefore important to assess cultural variation without assuming that everyone within a culture is essentially the same, and without putting undue emphasis on either the between-population differences or similarities.

Extending the study of leadership to include evolutionary theory and data from diverse, small-scale and traditional societies (i.e., non-WEIRD) can facilitate more generalizable theories and offer applications to shape organizational structure across populations (i.e., WEIRD settings) in ways which are more consistent and less mismatched with human evolutionary history and evolved leader-follower psychologies [@von_rueden_leadership_2015; @garfield_review_2019; @van_vugt_selected_2010]. Such approaches can also incorporate the anthropological insight that diversity of cultural, social, and ecological contexts are strong drivers of human behavior and group dynamics [@wiessner_risk_1982; @henrich_big_2015].

<!-- as well as the costs and benefits to individuals involved. Cultural norms, capacities of competing leaders, inter-group dynamics, and follower preferences can also impact leader agendas. The interactions between leader strategies, pressures of followership, and group dynamics are likely to strongly contribute to organizational variation across human populations including social and economic trajectories  and cultural and biological evolutionary processes [].   -->

@@

Evolutionary theoretical models and empirical studies often seek to integrate findings within a broad framework of human cultural diversity. The universality of human leadership systems and the incredible diversity among the qualities of leaders across cultures and contexts and the functions leaders serve is difficult to capture. Scholars are then left to haphazardly self-select ethnographic cases or rely on the self-selected reviews of others for generalizations [e.g., @garfield_review_2019; @von_rueden_leadership_2015]. Here, we seek to provide the first empirical and near comprehensive view of the qualities and functions of leaders across human societies from a representative cross-cultural sample of "non-WEIRD" populations. We also provide the first comparative view describing the costs and benefits of leadership for both leaders and followers in an evolutionary biological framework. We also aim to explore how the qualities and functions of leaders vary within populations across social contexts and across populations with different subsistence strategies.

In sum, we sought to empirically and systematically accomplish the following four goals: 1) identify variation in the *qualities of leaders* and the *functions they serve*. 2) Identify *features* of the qualities and functions of leaders (e.g., do particular qualities or functions tend to covary or cluster together?) in reference to theoretical models or novel interpretations. Such features might lend support to existing theories; for instance they might correspond to qualities associated with "prestige" or "dominance" [@henrich_evolution_2001] or functions associated with conflict or cooperation [@gavrilets_collective_2015; @glowacki_evolutionary_2017; @hooper_theory_2010]. 3) Identify variation in the *costs and benefits* of leadership for both leaders and followers. And 4) determine the extent to which the qualities and functions of leaders, and their costs and benefits, varied across social contexts within populations, varied across popululations with different subsistence strategies, and varied by continental region.

Our approach was exploratory: we did not have *a priori* hypotheses to test.

<!-- Such exploratory results stand to facilitate more fine-grained approaches within leadership studies and greater theoretical and empirical work beyond community leadership, while also revealing the importance of a broad, cross-cultural perspective in understanding the psychology and cultural patterning of perhaps the most foundational component of human social behavior and organization. -->

@@

We estimated the cross-cultural frequency in the functions and qualities of leaders and the costs and benefits of leadership for both leaders and followers using an ethnographic database previously built and published (doi.org/10.5281/zenodo.2541999), from which @garfield_evolutionary_2019 tested three theoretical models against the ethnographic record. The electronic Human Relations Area Files (eHRAF) is a digital collection of ethnographies, subject-coded at the paragraph level following the Outline of Cultural Materials (OCM) coding scheme. The eHRAF is a subset of the larger HRAF and is continually growing in scope.

Whereas @garfield_evolutionary_2019 documented supporting evidence for measures of leadership in reference to influential theoretical models, our goals here are to let the ethnographic record "speak for itself" in understanding variation in the dimensions of leadership we are interested in, i.e., our methods and analyses are exploratory and not strictly theoretically driven. We draw on the same database of `r nrow(all_data)` ethnographic paragraphs (termed text records) extracted from the 60-culture Probability Sample Files (PSF) of the eHRAF database [@naroll_proposed_1967]. The PSF aims to mitigate Galton's problem by randomly sampling of one culture with high quality ethnographic coverage from each of 60 culture regions. It comprises over `r trunc(sum(leader_cult['ehraf_pages'])/1000)`K pages of digitized text.

@@

The operationalized qualities and functions (See Table \@ref(tab:qualfuncvars)) were developed from the ethnographic text records in the database. First ZG read a subset of text records and developed initial variable lists and operationalizations of the qualities and functions in those text records. Then, KS coded approximately half the `r nrow(all_data)` text records (in a random order) for the operationalized qualities, functions, cost, and benefits, while also creating new variables and operationalizations for additional qualities and functions encountered. After coding half of the random sample, KS and ZG revised the operational definitions to produce the final list of variables in Table \@ref(tab:qualfuncvars). KS and ZG then coded the entire sample of `r nrow(all_data)` text records for the `r length(c(quality_vars, function_vars, leader_cost_vars, leader_benefit_vars, follower_benefit_vars, follower_cost_vars))` operationalized variables.

```{r qualfuncvars, echo=FALSE, message=F, warning=FALSE}
qual_func_var_tab<-read_csv('func_qual_definitions.csv', na = 'NA')
kable(qual_func_var_tab, caption = "Operational definitions for coded leader qualities and functions.")
```

In constrast to the qualities and functions variables, coding operationalizations of the costs and benefits of leadership were developed *a priori* based on our reading of the evolutionary theoretical literature from anthropology and biology. For both followers and leaders, we coded if there was evidence of costs paid or benefits received in the following categories: social status, reduced or increased risk of harm or conflict, social services, territory, mating, fitness, food resources, material resources, and other miscellaneous costs or benefits not captured by these categories. See Table \@ref(tab:costbenefitvars). There are multiple instances of conceptual overlap between many of our researcher-coded variables, in particular between various qualities and related costs or benefits. The operationalization of costs and benefits generally capture a wider range of evidence than related qualities and focus more on evidence of transaction rather than description. We return to this point in the limitations section.

```{r costbenefitvars, echo=FALSE, message=F, warning=FALSE}
cost_benefit_vars_tab<-read_csv('costs_benefits.csv', na = 'NA')
kable(cost_benefit_vars_tab, caption = "Operational definitions for coded benefits and costs for leaders and followers.")
```

ZG then created a measure of group structure coded as: *residential subgroup*, which includes informal groups of co-residents, age-based groups, performance groups; *kin group*, which includes lineages, phraties, and other kin-based organizations; *economic group*, which includes subsistence groups, market groups, and other groups with primarily economic goals; *military group*, which includes all groups related to inter-group conflict; *religious group*, which includes groups collected for spiritual or supernatural purpose; *political group (community)*, which includes political groups at the level of the community, i.e., the political leader can potentially directly interact with most community members such as most village headmen, or *political group (supracommunity)*, which includes political groups that encompass multiple residential communities, such as complex chiefdoms, regional political leaders, and kings or state-level leaders. ZG and KS then coded all text records for group structure.

ZG and KS then discussed every instance of disagreement between their respective coding on all variables to produce a consensus matrix, which is used in the analyses here and added to the published data repository (in a data frame named *leader_text2*). We illustrate our coding with an example from Andamanese hunter-gatherers [@az02-001, p. 45]:

@@

This text record was coded has having evidence for the functions *Resolve conflict*, *Organize cooperation*, *Movement/migration*, *Provide subsistence*, and *Misc. social functions*. For qualities, there was evidence for *High status*. Although authority is mentioned, we did not code evidence for this variable because the chief "possesses no power to punish or enforce obedience to his wishes," contrary to our operationalization.

Each culture was coded for subsistence strategy. We rely on a modified classification of subsistence from the HRAF, collapsing “hunter-gatherers” and “primarily hunter-gatherers” into a single category, “hunter-gatherers,” and by collapsing “agro-pastoralists” and “pastoralists” into a single “pastoralists” category. We renamed the HRAF’s “Other” subsistence category, which is applied to societies with various combinations of hunting, fishing, gathering, pastoralism, and horticulture or intensive agriculture, as “mixed.” This produced a subsistence classification with five groups: Hunter-gatherers, Pastoralists, Horticulturalists, Agriculturalists, and Mixed.

For additional details on ethnographic sampling and creation of the *leadershipdata* package see @garfield_evolutionary_2019.

@@

Our data comprised one row for each of our text records. For each of our variables, each row was coded as 1 if there was evidence in the text record for that variable, and 0 otherwise. Because absence of evidence is not evidence of absence, our data only inform the extent to which there is evidence for each of the operationalized leader traits.

We analyzed data from our researcher-coded operationalized variables at the text-record level and at the culture-level. Our first goal was to assess the proportion of text records and cultures that provided evidence for each of our coded variables to discover the qualities, functions, benefits and costs of leadership that are most and least frequently noted in the ethnographic record. We computed the proportion of text records supporting each variable, and their 95% confidence intervals, using logistic generalized linear mixed effects models with random intercepts for document authors nested within cultures (which partially accounts for potential variation due to differences in ethnographers such as gender, nationality, or theoretical framework), controlling for one potential source of bias as a fixed effect: the date of publication. For each varible, we then computed the proportion of cultures with at least one text record with evidence for that variable, with confidence intervals computed using a cluster bootstrap.

@@

Our third goal was to determine which leader and follower benefits and costs were most commonly mentioned in the ethnographic record, computing proportions and 95% CIs as above and to estimate how costs and benefits varied by group structure types.

@@

The geographic distribution of the cultures in our sample is displayed in Figure \@ref(fig:geographic), with cultures labeled by subsistence strategy and represented by a symbol proportional to the number of documents that provided ethnographic texts for that culture (see also Figure \@ref(fig:subsisttree)). Our search yielded `r nrow(all_data)` text records from `r length(unique(leadershipdata::documents$d_ID))` unique documents describing `r length(unique(all_data$d_culture))` cultures. The number of text records per culture ranged from `r min(leader_cult$number_leader_records, na.rm=T)` to `r max(leader_cult$number_leader_records, na.rm=T)`, with a median of `r median(leader_cult$number_leader_records, na.rm=T)`, a mean of `r signif(mean(leader_cult$number_leader_records, na.rm=T), 3)`, and a standard deviation of `r signif(sd(leader_cult$number_leader_records, na.rm=T), 3)`. The text records were generally short, ranging from `r textstats$min` to `r textstats$max` words, with a median word count of `r textstats$median`, a mean of `r textstats$mean`, and standard deviation of `r textstats$sd`. Of the `r nrow(all_data)` text records, `r final_record_count` had evidence for at least one variable in the study.

@@

@@

Susistence strategy was confounded with continental region. African text records were predominantly from horticultural societies, for instance, whereas North American text records were prodominantly from hunter-gatherer societies (see Figure \@ref(fig:regionsubsistence)).

Susistence strategy and group structure were also confounded. Hunter-gatherer text records often refered to residential leadership (`r hg_residential_pct`% of the total sample of hunter-gatherer text records) and kin leadership (`r hg_kin_pct`%), but rarely refered to supracommunity leadership (`r hg_supra_pct`%). Agricultural text records, on the other handj, rarely referring to residental leadership (`r agri_residential_pct`%) but often referring to supracommunity leadership (`r agri_supra_pct`%). Across all subsistence types, the most frequent group type was political groups at the community level, with the exception of agriculturalists in which supracommunity political groups are the most frequent group type. Leadership beyond the community level is relatively rare outside of agricultural populations and is especially rare among hunter-gatherers. See Figure \@ref(fig:grouptype).

@@

## Leader sex

@@

@@

The number of text records on leadership retrieved for a particular culture was weakly correlated with the total number of pages of ethnography in the eHRAF (r = `r signif(cor_pages_tr$estimate, 2)`, p = `r signif(cor_pages_tr$p.value, 2)`). A loess regression suggested that the number of text records initially increased with the total number of pages of ethnography, plateauing at about 35 records around 4500 pages (see Figure \@ref(fig:eHRAFpages)).

@@

For each of the `r length(c(quality_vars, function_vars, leader_cost_vars, leader_benefit_vars, follower_benefit_vars, follower_cost_vars))` coded variables (see Tables \@ref(tab:qualfuncvars) and \@ref(tab:costbenefitvars)), we computed the proportion of text records (95% CI) that provided support using logistic mixed effects models that controlled for one possible source of bias -- pubication date -- as a fixed effect, and included authors nested within cultures as random intercepts. We then computed the proportion of cultures with at least one text record providing support for that variable (calculating 95% CI using a cluster bootstrap). See Figure \@ref(fig:varsupport). For a table of all coefficients and 95% CI's, see Supplementary Table S??.

@@

We then investigated if the proportion of text records with evidence for high status differed among the different levels of social and political organization, subsistence strategies, continental regions, and leader sex (again using a logistic mixed effects model with these variables as fixed effects and authors nested within cultures as random intercepts). There were modest differences. Hunter-gatherers and pastorilist societies were more likely to provide evidence for high status, as was South America and East Asia. Female leaders were also more likely to be identified as high status. Somewhat counter-intuitively, the proportion of text records indicating high status was lowest at the highest level of political organization, perhaps because it would have been superfluous to note that kings and chiefs were high status. See Figure \@ref(fig:groupstatus). For the table of coefficients, see Supplementary Table X.

```{r groupstatus, fig.width=10, fig.cap="Estimated marginal means of a mixed effects logistic regression model of evidence for high status (0/1) as a function of group type, subsistence strategy, continental region, and leader sex, with random effects for ethnographer nested within culture. Values are on the response scale (probability)."}
((plot_emm_status_group2 / plot_emm_status_sex2) | (plot_emm_status_subsistence2 / plot_emm_status_region2)) + plot_annotation(title = "Predictors of high status", tag_levels = 'A')
```
